---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  word_document: default
  pdf_document: default
---


## Loading and preprocessing the data

```{r, echo = TRUE}
library(dplyr)
library(tibble)
library(ggplot2)

activity <- read.csv("activity.csv")



```

## What is mean total number of steps taken per day?

1.Calculate the total number of steps taken per day

```{r, echo=TRUE}
act2<-mutate_at(activity, "date", as.character)

act3<-mutate_at(act2,"date",as.Date,format = "%Y-%m-%d")

act4<-act3 %>% group_by(date) %>%  summarise(avg.steps.per.day=sum(steps))

head(act4)
```

2.Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}
hist(act4$avg.steps.per.day, xlab = "Average steps per day", ylab = "Frequency",main="",breaks = 10)

```

3.Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}
mean(act4$avg.steps.per.day, na.rm = TRUE)

median(act4$avg.steps.per.day, na.rm = TRUE)
```

## What is the average daily activity pattern?

1.Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}

act5<-act3 %>% group_by(interval) %>%  summarise(avg.steps.per.interval=mean(steps,na.rm = TRUE))

plot(act5$interval,act5$avg.steps.per.interval,type = "l",xlab = "Interval", ylab = "Average steps per day",main="")

```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
order(act5$avg.steps.per.interval, decreasing = TRUE)

act5[104,]
```


## Imputing missing values

1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo=TRUE}
summary(is.na(act3$steps))

```

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}

act6<-full_join(act3,act5,by = "interval")

new.steps<-numeric(0) 

for (i in 1:17568){
  if (is.na(act6$steps[i])) { 
    new.steps<-append(new.steps,act6$avg.steps.per.interval[i])
  } else {
    new.steps<-append(new.steps,act6$steps[i])
  }
}

act7<-cbind(act6,new.steps) 


act8<-as.tibble(act7)

head(act8)
```


4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r, echo=TRUE}
act9<-act8 %>% group_by(date) %>%  summarise(new.avg.steps.per.day=sum(new.steps))

par(mfrow = c(1, 2))
hist(act9$new.avg.steps.per.day,xlab = "New average steps per day", ylab = "Frequency",main="",ylim = c(0,25),breaks = 10)
hist(act4$avg.steps.per.day,xlab = "Average steps per day",main="",ylim = c(0,25),breaks=10)
```

Mean without missing values
```{r, echo=TRUE}

mean(act9$new.avg.steps.per.day)
```

Median without missing values
```{r, echo=TRUE}

median(act9$new.avg.steps.per.day)
```

Mean with missing values
```{r, echo=TRUE}

mean(act4$avg.steps.per.day, na.rm = TRUE)

```

Median with missing values

```{r, echo=TRUE}

median(act4$avg.steps.per.day, na.rm = TRUE)
```

Based on the mean and median values (for all days) for the average steps taken per day, their is not much difference in the original data with missing values compared to the new dataset with missing values replaced.  However, based on the plotted histogram of the 2 datasets, there is cleary a difference.  A more in depth analysis should be performed to investigate this.

## Are there differences in activity patterns between weekdays and weekends?

1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}

library(timeDate)
act11<-act8%>%mutate(weekend.or.not=isWeekday(date,wday = 1:5))   
head(act11)

```

2.Make a panel plot containing a time series plot (i.e.
type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}

act11<-act8%>%mutate(weekend.or.not=isWeekday(date,wday = 1:5))   

act12<-mutate_at(act11, "weekend.or.not", as.character)

act13<-mutate_at(act12, "weekend.or.not", as.factor)

act14<-mutate_at(act13, "weekend.or.not", gsub,pattern="FALSE",replacement="weekend")
act15<-mutate_at(act14, "weekend.or.not", gsub,pattern="TRUE",replacement="weekday")

act16<-act15 %>% group_by(weekend.or.not,interval) %>%  summarise(avg.steps.per.weekday.or.wkd=mean(new.steps))

ggplot(act16, aes(interval, avg.steps.per.weekday.or.wkd, fill = weekend.or.not))+
  geom_line(aes(color= weekend.or.not))


```





